Continous time Markov chain models; \\ extension, simulation and visualisation

### 2015-02-10 Response.tex

% consistent edits between manuscript and rebuttal letter
\input{thesis.xtr}

\setcounter{page}{1}

\bigskip
\medskip
\noindent{\Large 
Response 
}
\bigskip

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%---SUGGESTED SIMULATION SCENARIOS---%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \section*{Simulations}
% 
% \begin{itemize}
% 
% \subsection*{David Posada}
% 
% \item {
% {\it
% \textbf{
% Simulate a number of epochs that do not correspond with the assumed ones.
% }% END: textbf
% }% END: it
% }% END: item
% \\
% Running. \\
% Simulate 3, infer 2 : simulationSet2 \\
% Simulate 2 infer 3 : simulationSet1 
% 
% 
% \item {
% {\it
% \textbf{
% Simulating under a non-epoch model, for example where different process can apply to contemporaneous branches, would be fundamental to understand the robustness of the model.
% }% END: textbf
% }% END: it
% }% END: item
% \\
% Running (trend\_simulation).
% 
% 
% 
% \item {
% {\it
% \textbf{
% Here it would have been interesting to see the correspondence between simulated and estimated tMRCAs when the true model has one epoch with omega=0.5 (or even better, omega equal to the weighted mean of the n-epoch model).
% }% END: textbf
% }% END: it
% }% END: item
% \\
% Running (deepRootv2).

@@

\subsection*{Stijn Vansteelandt}

\begin{itemize}

\item {
{\it
\textbf{
It would also have been nice to understand better whether the use of the
epoch model (versus not) has an essential impact on tree reconstruction.
}% END: textbf
}% END: it
}% END: item
\\
We are performing a set of simulations to asses this, and the obrained results will be presented during the public defence.
% Running (epoch\_tree\_simulation1 epoch\_tree\_simulation2)

\item {
{\it
\textbf{
How mislead one could be if the data were generated under an epoch model, but analyzed under a simpler model.
}% END: textbf
}% END: it
}% END: item
\\
We are performing a set of simulations to asses this, and the obrained results will be presented during the public defence.

@@

\begin{itemize}

% 1
\item {
{\it
\textbf{
The thesis starts with a very nice introduction, which I think is very insightful for the reader who is not familiar with phylogenetic and/or Bayesian inference. 
In spite of this, I believe that the readability of the later chapters would be further improved if the introduction:
}% END: textbf
}% END: it
}% END: item

\begin{itemize}

% 1a
\item {
{\it
\textbf{
Briefly described the evolutionary models, such as GY94, that are later used.
}% END: textbf
}% END: it
}% END: item
\\
Corrected, there is now a \emph{Markov models of codon evolution} subsection in the introductory chapter, which discusses most popular codon model implementations (see Subsection ~\ref{sub:codon}).

% 1b
\item {
{\it
\textbf{
Briefly described how to accommodate time-stamped sequences, which is now considered in the analyses in Chapter 2, but not really discussed.
}% END: textbf
}% END: it
}% END: item
\\
Corrected, see Subsection~\ref{sub:dated_trees}, which describes the methods used in BEAST to accomodate time-stamped sequences as a source of calibration for trees.

% 1c
\item {
{\it
\textbf{
Give a bit more detail on phylogeography, e.g. how do the models look like and what is the connection between phylogeography and epoch models (some connection is described in Chapter 2, but remains a bit vague).
}% END: textbf
}% END: it
}% END: item
\\
I add a figure and additional explanations in Subsection~\ref{sub:phylogeo}, which should make the subsequent chapters clearer now..

@@

\end{itemize}

% 2
\item {
{\it
\textbf{
It is sometimes unclear in Chapter 2 whether the analysis is performed under a fixed evolutionary tree or not. 
This is especially so in the evaluation of Bayes Factors in Table 2.3. 
It would also have been nice to understand better whether the use of the epoch model (versus not) has an essential impact on tree reconstruction. 
Somewhat related, it would have been nice to learn from the simulation study how mislead one could be if the data were generated under an epoch model, but analyzed under a simpler model. 
I am not asking for additional analyses here, but merely for additional clarification or insight.
}% END: textbf
}% END: it
}% END: item
\\
% FB: I don't get the first part of this comment. In the Intro we clearly write:
% "Our Bayesian approach also does not condition on a fixed tree topology but averages over all plausible evolutionary histories." (p 46) \\
% PL: OK, but to make it even more explicit I have added this to the results section (in simulation results): 'In all our analyses of the replicate data, as well as in the analyses of the real-world data sets in the next sections, we consider the tree as random and estimate it from the nucleotide data.' (this still needs to be tagged).
To re-iterate the fact that our model does not condition on a known tree topology we now write:

\begin{quote}
\myeditsvtwo
\end{quote}

We now also entertain a number of simulations which asses the impact of correcting for time-heterogeneity on tree reconstruction. 
The simulations which address the second part of this comment are now being performed, I will discuss them during the public defence.

% 3
\item {
{\it
\textbf{
P6, lines 5-7: this is not correctly phrased as the parameter values do not have an associated probability in the frequentist likelihood framework.
}% END: textbf
}% END: it
}% END: item
\\
This paragraph is now rephrased:

@@

\begin{quote}
\myeditsvthree
\end{quote}

% 4
\item {
{\it
\textbf{
Expression (1.12) is confusing in that indices i and j appear in the lefthand side, but not in the righthand side. 
Likewise in expression (1.13).
}% END: textbf
}% END: it
}% END: item
\\
Corrected.

% 5
\item {
{\it
\textbf{
P20, 2nd display: an equality sign is missing on the 2nd line.
}% END: textbf
}% END: it
}% END: item
\\
Corrected.

% 6
\item {
{\it
\textbf{
I would find it useful to see a brief discussion on the plausibility of (1.22).
}% END: textbf
}% END: it
}% END: item
\\
% FB: Does he mean plausibility from a biological perspective or statistical? It's a necessary condition for the existence of stationary dist.
% Pl: I see it as questioning the plausibility in a biological sense. We now already argue that it is an assumption taken for computational reasons, but at the end of that paragraph we can explicitly state 'Realistic substitution processes are however unlikely to be time-reversible and procedures have also been developed to compute phylogenetic likelihoods under non-reversible models (cite Boussau and Gouy, 2006, see below)'. ...perhaps add in something more from that paper.... and end with 'We note that we use such non-reversible models in our discrete phytogeographic applications (cite Edwards et al., Current Biology, 2011).
% http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.324.620&rep=rep1&type=pdf
Please see my changes to Subsection~\ref{sub:subst_models}.
We explain the implications of this assumption from biological perspective, as well as recent attempts to relax this assumption, both in the Bayesian phylogeographic framework and more bradly in th econtext of phylogenetic applications.

@@

% 7
\item {
{\it
\textbf{
If I am right that matrix P(t) on the bottom of page 22 is four-dimensional, then why not write it in full, which will be less confusing?
}% END: textbf
}% END: it
}% END: item
\\
Corrected.

% 8
\item {
{\it
\textbf{
P25, the statement at the end of the penultimate paragraph (`most plausible hypothesis for the data') is vague.
}% END: textbf
}% END: it
}% END: item
\\
This statement now reads:

\begin{quote}
\myeditsveight
\end{quote}

% 9
\item {
{\it
\textbf{
P28, I understand what T(n) is, but would nevertheless find it helpful if it were stated more explicitly.
}% END: textbf
}% END: it
}% END: item
\\
% FB: I don't understand the question in the definition we clearly define T as being simply a function on a positive subset of real numbers?
% Does he mean we should wrote "time"? but this is not always true?
% PL: OK, but can we then state something like this: T(n) can be interpreted as time in most practical situations, but this does necessarily need to be the case, e.g.....?
Corrected, we now add a passage which explains that we mean the ``worst case time'' scenario.

@@

% 10
\item {
{\it
\textbf{
P31, If the phrase `proper prior' is used, then perhaps it should be defined.
}% END: textbf
}% END: it
}% END: item
\\
This paragraph now reads:

\begin{quote}
\myeditsvnine
\end{quote}

% 11
\item {
{\it
\textbf{
P32, Since the author assumes that the reader may have no prior knowledge of Bayesian inference, I believe the meaning of Figure 1.11 should be explained in more detail.
}% END: textbf
}% END: it
}% END: item
\\
Corrected. There is now a more elaborate description accompaning this figure (see Subsection~\ref{sub:ex_beta_binom}).

@@

\begin{quote}
\myeditsvfifteen
\end{quote}

% 16
\item {
{\it
\textbf{
P44. I would find it useful to receive a bit more guidance concerning the connection to phylogeographic inference. 
Also in Section 2.4.3, the description of the proposal by Bahl et al. (2011) is too brief for me to appreciate why the epoch model is more appropriate.
}% END: textbf
}% END: it
}% END: item
\\
% FB: This becomes p46 now, from "The former is critical to accommodate discrete phylogeographic inference"
I added a sentence taken from Bahl et al.:

\begin{quote}
\myeditsvsixteen
\end{quote}

This further clarifies the distinction between the \emph{ad-hoc} method by Bahel et al. and the ``epoch model'' applied to the global influenza spread problem.

% 17
\item {
{\it
\textbf{
Display (2.5): I believe this display implicitly assumes that the time T1 is a priori known. 
It would be good to state this explicitly.
}% END: textbf
}% END: it
}% END: item
\\
Corrected.

@@

% 18
\item {
{\it
\textbf{
P57. The sentence `To determine the posterior odds, ... of that model' is vague.
}% END: textbf
}% END: it
}% END: item
\\
The sentence is rephrased:

\begin{quote}
\myeditsveighteen
\end{quote}

% 19
\item {
{\it
\textbf{
P85. Riemannian -> Riemann
}% END: textbf
}% END: it
}% END: item
\\
Corrected.

% 20
\item {
{\it
\textbf{
P86. The last sentence of Section 5.1.3 is not sufficiently clear. 
Please make clear why the use of mixed effects models could allow this.
}% END: textbf
}% END: it
}% END: item
\\
I changed the Subsection~\ref{sub:glm_epoch}, explaining that by this sentence we mean the impact of various covariates as well as that by using MG94 parametrization we get a better resolution on selection parameters.
There is also a more elaborate description of the model we plan to use in the upcoming manuscript on the application of the hierarchical epoch model.

% 21
\item {
{\it
\textbf{
P88. It would be useful to receive a bit more detail what is meant by `infinite hidden Markov Model'.
}% END: textbf
}% END: it
}% END: item
\\
I introduce the definition of HMM's in a form of Remark, please see Subsection~\ref{sub:dpp}.

@@

\end{itemize}

%%%%%%%%%%%%%%%%%%%%
%--- COMMENTS---%
%%%%%%%%%%%%%%%%%%%%
\section*{Comments by David Posada}

\begin{itemize}

% 1
\item {
{\it
\textbf{
I missed a brief primer of Bayesian model selection in the introductory chapter.
Techniques such Bayes factors, path sampling or stepping-stone sampling are used throughout but never explained. 
This should be very easy to fix.
}% END: textbf
}% END: it
}% END: item
\\
There is now a subsection called \emph{Bayesian model selection} at the end of Chapter 1, which gives a primer on Bayesian model selection that should be sufficient to follow the rest of the manuscript.

% 2
\item {
{\it
\textbf{
In the simulations in Chapter 2 you assume you know the true epoch times, but this information is rarely available in real life. 
Simulation 1 addresses the impact of model misspecification partially, as the true process does not have epochs. 
It could be a good idea to further check the impact of model misspecification, in particular simulating a number of epochs that do not correspond with the assumed ones (you later mention this idea in the final chapter but do not end up testing its potential biasing effect). 
Also, simulating under a non-epoch model - for example where different process can apply to contemporaneous branches, would be fundamental to understand the robustness of the model. 
This does not need to be done before the defense, but it could be a nice extension for the future.
}% END: textbf
}% END: it
}% END: item
\\
We are performing a set of simulations to asses this, and the obrained results will be presented during the public defence.

% 3
\item {
{\it
\textbf{
In general, there is some repetitive text, sometimes with identical phrases/paragraphs, between the introduction and some of the other chapters. 
On the other hand, some relevant aspects of the research are discussed in the general discussion in Chapter 5, instead of in the corresponding `research' chapter. 
This is true in particular for Chapter 3, were a very concise description of SPREAD is provided. 
I understand these redundancies and fragmentations are a byproduct of reproducing papers as chapters, but however these issues have been corrected in some occasions. 
For example, chapter 2 is not identical to the SysBio paper. 
I suggest you try a bit to further consolidate the different chapters.
}% END: textbf
}% END: it
}% END: item
\\
We carefully screened for repeated paragraphs between the introduction and the other chapters, and could only identify one paragraph that is repeated at th ebeginning of Subsection~\ref{sub:subst_models}.
This paragraph is now re-sentenced to remove the repetition.
We acknowledge that isolated sentence might still carry some degree of resemblance, yet we preffer not to remove this in order to also allow the subsequent chapter to function as self-contained work, for readers possibly interested only in a part of the presented work.

% 4
\item {
{\it
\textbf{
The simulation software piBUSS described in Chapter 4 can be very useful, in particular in regards to its connectivity with BEAST. 
The possibility of simulating discrete phylogeographic traits is very convenient. 
Here it would have been interesting to see, for the sake of comparison and interpretation, the correspondence between simulated and estimated tMRCAs when the true model has one epoch with omega=0.5 (or even better, omega equal to the weighted mean of the n-epoch model). 
This does not need to be done before the defense, but it could be a nice extension for the future.
}% END: textbf
}% END: it
}% END: item
\\
We are performing a set of simulations to asses this, and the obrained results will be presented during the public defence.

% 5
\item {
{\it
\textbf{
Page 14: I would say that the term `Big Data' does not apply to sequence alignments. 
It might do for NGS files etc., but not for the type of data that BEAST uses as input. 
Not yet. 
Moreover, you focus on HPC architectures like GPU and multicores, which are in some senses opposite to Big Data architectures (i.e., Hadoop Map Reduce).
}% END: textbf
}% END: it
}% END: item
\\
I have rephrased the entire paragraph to avoid using the term:

@@

\begin{quote}
\myeditDPfive
\end{quote}

% 6
\item {
{\it
\textbf{
Page 25, 31: likelihood should be $P(X| \theta)$, not $P(\theta | X)$.
}% END: textbf
}% END: it
}% END: item
\\
Corrected.

% 7
\item {
{\it
\textbf{
Page 34: on the left-term of equation 1.37, the denominator should be
$P(\theta[t]|X) q(\theta^{*} | \theta[t])$
not 
$P(\theta[t]|X) q(\theta[t] | \theta^{*})$
}% END: textbf
}% END: it
}% END: item
\\
Corrected.

% 8
\item {
{\it
\textbf{
Page 52: according to the legend in Figure 2.3, a two-epoch plot is missed.
}% END: textbf
}% END: it
}% END: item
\\
Entire set up for the employed simulations is presente din on eplot by th euse of alternating shading. 
This was perhaps not clear enough from the Figures caption, I have thus rephrased it to state explicitely:

@@

% 10
\item {
{\it
\textbf{
Page 137: Figure B.12 should be in color. 
Otherwise it is very difficult to grasp.
}% END: textbf
}% END: it
}% END: item
\\
Corrected, all the figures presenting SPREAd output are now in color.

% 11
\item {
{\it
\textbf{
Page 140: Figure B.13 should be in color. 
Otherwise it is very difficult to grasp.
}% END: textbf
}% END: it
}% END: item
\\
Corrected, all the figures presenting SPREAd output are now in color.

% 12
\item {
{\it
\textbf{
Page 143: Figure B.15 should be in color. 
Otherwise it is very difficult to grasp.
}% END: textbf
}% END: it
}% END: item
\\
Corrected, all the figures presenting SPREAD output are now in color.

% 13
\item {
{\it
\textbf{
Unless I missed it, the appendices are not cited in the text.
}% END: textbf
}% END: it
}% END: item
\\
We now reffer to the Appendices in the Chapters to which they apply.

### 2015-02-10 chapters/chapter01.tex
implementing comments

% TODO markov property is inherited from Poisson process
Drawing realizations with probabilities defined by $\mathbf{P}$  gives rise to a continuous-time stochastic process $\left\{ X(t):\; t\geq0\right\}$ satisfying the Markov property, such that for every $n\geq 0$, given the time points $0\leq t_{0}\leq t_{1}<\ldots<t_{n}\leq t_{n+1}$ and discrete states $i_{0},i_{1}, \ldots, i_{n},i_{n+1}$, it holds that:

@@

\subsection{Example: Comparing two sequences with the beta-binomial model}

>>>>

After obtaining finite-time transition probabilities given by Equation~\ref{eq:eigen_decomposition} for some time $t\geq0$ one can start drawing realizations, which follow probabilities defined by $\mathbf{P}$.
These realizations constitute a continuous-time stochastic process $\left\{ X(t):\; t\geq0\right\}$, which inherits the Markov property from the Poisson process modelling the time between any two mutations, as defined in Subsection~\ref{sub:poisson}.

We say that a stochastic process has the Markov property if the conditional probability distribution of future states of the process depends only upon the present state, not on the sequence of events that preceded it, i.e. for every $n\geq 0$, given the time points $0\leq t_{0}\leq t_{1}<\ldots<t_{n}\leq t_{n+1}$ and discrete states $i_{0},i_{1}, \ldots, i_{n},i_{n+1}$, it holds that:

@@

\subsection{Example: Comparing two sequences with the beta-binomial model\label{sub:ex_beta_binom}}
